GPU-Beschleunigung der Material Point Method

### 2019-02-07 pres/pres.tex
pres begin

\documentclass[m,times]{cgMA}
\usepackage{listings}
\usepackage{xurl}
\urlstyle{sf}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{color}
\usepackage{subfigure}
\usepackage{wrapfig}
\usepackage[english,ngerman]{babel}
\usepackage[backend=biber,style=alphabetic]{biblatex}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{mathtools}
\usepackage{subdepth}
\usepackage{enumitem}
\usepackage[toc,page]{appendix}
\usepackage{minted}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{svg}
\usepackage{xcolor,colortbl}
\addbibresource{mscthesis.bib}

@@

\begin{document}
\author{Fabian Meyer}
\title{GPU-Beschleunigung der Material Point Method}
\zweitgutachter{Bastian Krayer, M.Sc.}
\zweitgutachterInfo{(Institut f{\"u}r Computervisualistik, AG Computergraphik)}
% Umschalten der Sprache (fuer englische Rubrikbezeichnungen etc.)

\pagenumbering{roman}
\maketitle

\clearpage
\selectlanguage{english}
\vfill
\begin{center}
\subsubsection*{Zusammenfassung}
\end{center}
Die Material Point Method hat sich in der Computergrafik für physikalische basier\-te Simulationen etabliert. Sie benutzt ein hybrides Modell aus Lagrange-Partikeln und Euler-Gitter. Während Partikel als konsistenter Speicher fungieren, erlaubt das Gitter anfallende partielle Differentialgleichungen effizient zu lösen.

\noindent Die Material Point Method unterliegt hohen Berechnungszeiten, die die Methode nur für Hero Shots rentabel macht. Sie ist jedoch hoch-parallelisierbar. Diese Arbeit zeigt, wie die Methode mit GPGPU-Techniken beschleunigt werden kann.

\noindent Der Datenaustausch von Partikel und Gitter wird über interpolierende Transfers erreicht. Pro physikalischen Zeitschritt werden diese mehrmals ausgeführt. Vorverarbeitungsschritte können unternommen werden, um diese Transfers performanter zu machen.

\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Cache Hitraten. Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen er\-mög\-lich\-en. Die Größen der Bins sind für keine Operation beschränkt. Weiterhin werden nur Blöcke ausgeführt, die Informationen erhalten.

\vfill
\begin{center}
\subsubsection*{Abstract}
\end{center}
The material point method is allowing for physically based simulations. It has found its way into computer graphics and since then rapidly expanded. The material point method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enables solving of various partial differential equations with ease.

\noindent The material point method suffers from high execution times and is thus only viable for hero shots. The method is however highly parallelizable. Thus, this thesis proposes how to accelerate the material point method using GPGPU techniques.

\noindent Core of the material point method are grid and particles transfers that interpolate between the two structures. These transfers are executed multiple times per physical time step. Preprocessing steps might be taken if their computing time is outweighed.

\noindent Deep sorting with counting sort increases coalescing and L2 cache hit rates. Binning allows to divide the grid into blocks for shared memory filtering techniques. All operations do not rely on fixed bin size. As another preprocessing step, only grid blocks are executed which have particles in them.
\vfill
\clearpage
\tableofcontents

%\setcounter{page}{3}
\clearpage         % oder \cleardoublepage bei zweiseitigem Druck
% \listoffigures   % fuer ein eventuelles Abbildungsverzeichnis
% \clearpage
\pagenumbering{arabic}

\section{Introduction} \label{intro}
General purpose computation on graphics processing unit(GPGPU) has elevated graphics processing units(GPUs) to any computational field utilizing massively parallelizable algorithms. As such is the nature of discretizations in physics.

The higher parallelization acceleration of GPUs is a blessing for those applications that can profit from it. In unison with efficient algorithms they still have much potential to offer. Conversely, algorithms which are not parallelizable perform much worse and should be avoided. Dividing up these tasks between CPU and GPU is not recommendable. Transferring data over the PCI-Bus between the two is slow for big amounts of data and requires synchronization. Tasks in such a pipeline should thus stay on the GPU although they might perform worse.

Programming for the GPU is another hurdle. GPU code is reliant on the generation and hardware vendor of the GPU to use relevant extensions. Backwards compatibility therefore is mostly ignored and written towards a certain GPU generation. This is amplified due to the GPU code often being reliant on architecture limits for maximum performance. The major languages to target the GPU are CUDA, OpenCL, Direct 3D, OpenGL, and Vulkan. They build on top of a C/C++-language subset.

Whereas CUDA offers the most libraries and optimized code, it is only available on NVIDIA GPUs. OpenCL's development on NVIDIA GPUs is dragging far behind the standard. Its merge into the Vulkan roadmap could be an incentive to support a more in-depth language universally \cite{OPENCL:ROADMAP}. Direct3D is only available on Windows. Vulkan is the successor of OpenGL and removes work from the driver to the developer. Main benefit over OpenGL is less CPU usage which helps little for the almost full GPU implementation here. This implementation opts for OpenGL as Vulkan is still relatively new. Everything presented could be implemented in one of the other languages as well.

The simulation of the dynamics of materials is still extremely challenging in different applications and fields. Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible. The material point method(MPM) has already reliable tackled many different models, materials, and even interaction between materials.

Pure Eulerian methods like the finite element method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability. Pure Langrangian methods form the end other of the spectrum. Prominent are smoothed-particle hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature. Particle in cell(PIC) techniques like the material point method are hybrid Lagrangian/Eulerian methods. They try to benefit from the nature of both. Advection on particles is trivial. Thus the material point method utilizes Lagrangian particles as a consistent storage of dynamic properties. The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid. The grid is unmoving and does not store any information between physical time steps.

Transfers between the Lagrangian and Eulerian are of need to bring the two together. These transfers are the focus of particle in cell techniques. These transfers are realized due to interpolation. Thus they create (estimable) numerical error and lead to dissipation. Moreover grid nodes are mostly outnumbered by particles leading to more numerical error as modes of the single particles are not realized by the grid.

The $p$ particles to one node relationship of a particle-to-grid(P2G)-transfer or inversely one node to $p$ particles of a grid-to-particle(G2P)-transfer is the main problem when realizing these operations on the GPU. When faced with the static architecture of the GPU, this will inevitably lead to branch divergence of threads within a subgroup as some nodes are assigned more particles than others.

Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as material point method extensions. In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out. The basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM to get an understanding of the prominent operations. Furthermore GPU intrinsics, metrics and optimization opportunities are discussed. Chapter \ref{sec:implementation} tends to the implementation. The evaluation of the implementation follows up in chapter \ref{sec:eval}. As a conclusion chapter \ref{sec:conclusion} gives a summarizing overview and tends to future work.

\clearpage
\section{Related Work}\label{sec:rel_work}
\textbf{Particle in cell} techniques were developed initially for hydrodynamics \cite{evans1957particle}. They are also widely used in plasma simulations to solve Maxwell's equations \cite{PIC:GPU}. Their key steps are identical to the MPM:
\begin{enumerate}
  \item Transfer particle data to a grid.
\item \label{it:solve_gov_eq}Solve the governing partial differential equations on the grid to move forward in time.
  \item Transfer grid data back to particles.
\end{enumerate}
In general the grid is regarded as a scratch-pad since its only a temporary storage for the means of computing item \ref{it:solve_gov_eq}. I.e. the grid could be deleted each physical times step. Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle(FLIP) was developed for fluid simulations and is still widely in use for large simulations. Their counterpart being the empirical smoothed-particle hydrodynamics \cite{gingold1977smoothed}.

As a further development the Affine-particle in cell(APIC) technique was introduced as an improvement to PIC and FLIP solvers. Due to the definition of a local affine velocity field around each particle (instead of a single vector) APIC conserves angular momentum, reduces damping and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}. \cite{MPM:APIC} \cite{MPM:OLD_APIC}

A step further goes PolyPIC giving the velocity field even more freedom by adding polynomial velocity modes of increasing order to the description. \cite{MPM:POLYPIC}

The \textbf{material point method} is a further development to the PIC methods. Both use analysis and numerics to be derived without the need for empirical assumptions. Originally it was developed by \cite{sulsky1995application} with a focus on the dynamics of solids. \cite{MPM:SNOW} brought the method into computer graphics for their animation film \textit{Frozen}. In rapid succession a variety materials and effects got modeled by the MPM. Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature, its fast rate of expenditure is not a surprise.

The simulation of (hyper-)elastic materials can be done with a number of different strain measures utilizing fixed Linear strain (chapter \ref{sec:cor_hyper}), Green-Lagrangian strain tensor (chapter \ref{sec:svd}), or logarithmic Hencky-strain (\cite{MPM:SHELLS}) for the different requirements of small to large strains. Inserting these into the different energy density $\boldsymbol{\Psi}(\boldsymbol{\epsilon})$ models can already result in a constitutive model. Plasticity in general requires extra modeling in terms of a flow rule, yield-condition, and a hardening rule which are not discussed here \cite{ochsner2014elasto}. To give a short impression of the developments in computer graphics:

\cite{MPM:PHASE_CHANGE} model phase transition (melting) to a liquid with using the fixed corotational hyperelasticity model (chapter \ref{sec:cor_hyper}) in combination with a discretization of the heat equation. \cite{MPM:OLROYDB} use an Olroyd-B model with Cauchy-Green strain to handle elasticity and plasticity for viscoelastic fluids, foams, and sponges. \cite{MPM:DRUCKER} uses Hencky-strain in combination with a St. Venant-Kirchhoff model and Drucker-Prager elastoplascity to model sand.
\cite{MPM:MULTI} combine particles of different species (sand and water) by one grid for each species and relating them with a heuristic momentum exchange. \cite{MPM:SHELLS} simulates thin shells (cloths,cups etc.) with a Kirchhoff-Love model using Hencky-strains.

There exist several implementations utilizing the \textbf{GPU for particle in cell} techniques, which the material point method is a part of:

\textbf{Particle sorting:} \cite{PIC:GPU} studies different particle list sorting methods. Methods like 'Message Passing Sort' \cite{kong2011particle} \cite{decyk2011adaptable}, 'In Place Particle-QuickSort' \cite{stantchev2008fast} and 'Linked List Ordering' \cite{burau2010picongpu} try to utilize the partial sorting and only update particles that need to be moved within the list. These methods often rely on the assumption that only a small number of particles change their grid node, where in contrast MPM can be quite dynamic. Those that do move, move at most to one neighboring grid-node or stay in the same node due to the CFL condition, chapter \ref{sec:cfl}. These methods often use fixed bin-sizes which can be quite memory intensive and add additional memory management. Fixed size binning makes the methods mentioned before easier to implement which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses. As a result \cite{PIC:GPU} come to the conclusion to fall back to the CUDA library \texttt{thrust}, although they do note that the aforementioned methods can perform faster for particles that move only a limited amount of grid nodes. \texttt{thrust} provides the radix sort method \texttt{thrust::sort\_by\_key()}, where one sorts by grid index as key.

In general deep or index sorts can be done. A deep sort reorders every variable of a particle. An index sort gives back an index with an offset to access the particle variable.

\cite{NVIDIA:NNSEARCH} compare radix to counting sort utilizing a uniform grid and use the latter for SPH-fluids.

This implementation compares index and deep counting sort and uses the latter. Although possible, this counting sort stays away from the limited movement implied by the Courant–Friedrichs–Lewy(CFL) condition. Fixed size bins are an assumption this implementation also does not want to impose on the method.

\textbf{Particle-to-grid-transfers} are generally done in two manners \cite{stantchev2008fast}:
\begin{enumerate}
  \item \textbf{Particle push:} X particles in a local neighborhood of a grid point push their attributes onto it. Since this relationship results in a race condition \texttt{atomicAdd(float f)}s or explicit synchronization is necessary, e.g. a\-tomic: \cite{PIC:GPU}, synchronized: \cite{FRANCESCO:ROSSI}.
  \item \textbf{Particle pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
\end{enumerate}
A \textbf{grid-to-particle-transfer} is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood. The grid points in a uniform grid are 'sorted' by nature. This implementation tests one particle pull and two particle push methods.

Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition. Shared memory is however limited per GPU (GTX 970, 48KB per block \cite{NVIDIA:GTX970}). The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block. Since the number of particles per grid point in the MPM is generally between 4-10, only very limited block sizes are available for pull methods that greedily load all particles \cite{MPM:SNOW}.

Stencil computations and filtering techniques underlie a very similar process to MPM-transfers. Although these techniques employ a one-to-one grid relationship, since no particles are involved, it is worth taking a look at them.
An interesting prospect for filtering techniques on the GPU is given by NVI\-DIA's new shuffle operations. Shuffle allows to 'share' register memory between threads of a warp \cite{NVIDIA:SHUFFLE}. They are however not supported on other architecture, yet.

Widely used is a 2.5D blocking: A domain (of the domain decomposition) is split into 2D-planes along one axis. Data reuse of shared memory is maximized by a cyclic-queue strategy: Start a thread for each grid node in the first plane and load the halo along the axis into shared memory. Then each thread iterates along the axis over the 2D-planes synchronously, discarding the oldest plane in memory while loading the next plane (along the axis) into shared memory. Thus there are no z-dimension halos. This however requires all particles in the halo to be present in the limited shared memory. As mentioned, this implementation wants to stay away from assumptions limiting particles similar to fixed bin sizes. \cite{brandvik2010sblock} \cite{williams2007scientific} \cite{krotkiewski2013efficient}

In particle push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
\begin{enumerate}
\item Atomically add up the values of the halo. This results in two atomic adds on a side node, four on an edge node, and eight on a corner node; If the block extend is bigger than half the halo extend, otherwise more.
\item Every halo of every block corresponds to actual global memory. After a transfer, do add steps along each of the three axes. This is infeasible for large supports due to the added memory requirements. \cite{crassin2011interactive}
\end{enumerate}

\textbf{Particle activation:} If the particle's velocity is under a user-defined threshold, it becomes deactivated. It does not get recognized by any particle-to-grid-transfer, nor the grid solve(, nor any grid-to-particle-transfer not affected by the velocity transfer). They may of course be reactivated if the velocity in the neighborhood exceeds that threshold again or they collide. \cite{MPM:GPU}

Dividing the particles up can be realized with stream compaction algorithms discussed in chapter \ref{sec:implementation}.

\textbf{Particle resampling:} A method that could alleviate pressure off fixed bin sizes and make shared-memory restrictions less of a factor is particle resampling. Their intention mostly is to fill numerical gaps in the material at large deformations where the material is not supposed to break yet. Split and merge methods can directly control how many particles in a cell are allowed and accordingly increase or reduce them.

A simple \textit{Split} method divides one particle into eight with a distance of half the diagonal length of a grid cell from the original particle. The eight particles span a cube that can be randomly rotated if desired. Mass and volume get equally distributed; Velocity and deformation gradient get copied. Refer to chapter \ref{sec:mat_point} or for conservation of mass and momentum in MPM. \cite{gao2017adaptive}

A \textit{Merge} method looks for the closest neighbors and creates the new merged particle at the geometric center. Mass and volume are accumulated. Velocity is computed from a mass-weighted average of the participating particles. The deformation gradient is accordingly to chapter \ref{sec:svd} decomposed into $\boldsymbol{F} = \boldsymbol{U}\boldsymbol{\Sigma}\boldsymbol{V}^T$. The quaternion average of $\boldsymbol{U}$s and $\boldsymbol{V}$s of the particles lead to an average rotation $\overline{\boldsymbol{U}}$ and $\overline{\boldsymbol{V}}$, respectively. The principal strains within $\boldsymbol{\Sigma}$ are just averaged component-wise to $\overline{\boldsymbol{\Sigma}}$. One can compose them in the same fashion leading to an average deformation gradient with $\overline{\boldsymbol{F}} = \overline{\boldsymbol{U}}\overline{\boldsymbol{\Sigma}}\overline{\boldsymbol{V}}^T$.\cite{gao2017adaptive}

Unconditionally keeping particles on a fixed size is not desirable due to fracture requiring that gaps will be created. \cite{gao2017adaptive} produce a signed distance field to identify interior points. Surface points also should not merge or split due to visual artifacts. Furthermore, the signed distance field can be used to adapt the grid size depending on the distance to the surface. This allows for coarser levels with increasing distance from the surface.
%look again for MPM/PIC gpu impl
%SPH, FEM methods
\clearpage
\section{Notation}\label{sec:notation}
Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$. Vectors furthermore use lower-case letters $\boldsymbol{v}$ with the exception of angular momentum $\boldsymbol{L}$; Matrices use upper-case letters $\boldsymbol{M}$. A variable in the Lagrangian form has a left subscript $0$: $ _0{\boldsymbol{x}}$. The Eulerian description has a left subscript $t$: $_t\boldsymbol{x}$. Occasionally, notation may be omitted where it is apparent from context.

Beginning from chapter \ref{sec:mat_point} variables defined on a particle will have right subscript $p$ (e.g. $\boldsymbol{x}_p$). Grid cells will be assigned a right subscript $i$ or additionally $j$ denoting the grid index (e.g. $\boldsymbol{x}_i$). This is not to be confused with the following Einstein-notation. Therefore, beginning with chapter \ref{sec:mat_point} Einstein-notation for components will use greek letters ($\alpha,\beta$). Occasionally summation over grid index $i$ and particles $p$ may be implied. A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n-1} \Delta t^i$ will have a right superscript: $\boldsymbol{x}^n$.

Used throughout this thesis is at instances the \textbf{Einstein-notation} when using tensors and vectors.

Repeated indices, that are also defined on the variable, imply component-wise operations. Let $\boldsymbol{a}$ and $\boldsymbol{b}$ be vectors of dimension $n$ and  $\boldsymbol{A,B,D}$ $ { m \times n }$ tensors. $a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $\boldsymbol{a}$ and $\boldsymbol{b}$. Vector and tensor/matrix addition thus become:
\begin{equation} \label{EINSTEIN:ADD}
  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad D _ { i j } = A _ { i j } + B _ { i j }.
\end{equation}
A transpose operator swaps the indices:
\begin{equation}
  D^T _ { i j }  =D_{ji}.
\end{equation}
Repeated indices, that are not otherwise defined, however imply summation on that index. The vector dot product becomes:
\begin{equation}
  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }.
\end{equation}
Following this notation the Frobenius inner product between two second-order tensors is (also called Frobenius scalar product):
\begin{equation} \label{EINSTEIN:FROBENIUS1}
  \boldsymbol { A } : \boldsymbol { B } \equiv A _ { i j } B _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } A _ { ij } B _ { ij }.
\end{equation}
The Frobenius inner product between a ${r \times s \times m \times n }$ fourth-order tensor $\boldsymbol{C}$  and second-order tensor creates a second order tensor combining the ideas of \ref{EINSTEIN:ADD} and \ref{EINSTEIN:FROBENIUS1}:
\begin{equation}
  \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }.
\end{equation}
Matrix-Vector and Matrix-Matrix multiplication can be expressed this way as:
\begin{equation}
  b_i = A_{i j} a_j \quad D _ { i j } = A _ { i k } B _ { k j }.
\end{equation}
The definition of the Kronecker Delta is as follows:
\begin{equation}
  \delta _ { i j } = \left\{
  \begin{array} { l l l }
      { 1 } & {\text{if}} &  {i = j}  \\
      { 0 } & {\text{if}} &  { i \neq j }
  \end{array}. \right
\end{equation}
The Kronecker Delta as an operator is most efficiently described as a substitution:
\begin{equation}
  a _ { i } \delta _ { i j } = a _ { j }.
\end{equation}
I.e. this component is only evaluated if $i = j$. The Kronecker Delta comes up in differentiating a variable by itself with same or different index:
\begin{equation}
  \frac{\partial{x_0}}{\partial{x_1}}=0, \frac{\partial{x_0}}{\partial{x_0}}=1  \Rightarrow \frac{\partial{x_i}}{\partial{x_j}} = \delta_{ij}.
\end{equation}
The alternating tensor or Levi-Civita symbol
\begin{equation}
  \varepsilon _ { i j k } = \left\{ \begin{array} { r l } { + 1 } & { \text { if } ( i , j , k ) \text { is } ( 1,2,3 ) , ( 2,3,1 ) , \text { or } ( 3,1,2 ) } \\ { - 1 } & { \text { if } ( i , j , k ) \text { is } ( 3,2,1 ) , ( 1,3,2 ) , \text { or } ( 2,1,3 ) } \\ { 0 } & { \text { if } i = j , \text { or } j = k , \text { or } k = i } \end{array} \right.
\end{equation}
is used to express cross-products $\boldsymbol{c}=\boldsymbol{a} \times \boldsymbol{b}$:
\begin{equation}
  c_i = \varepsilon_{ijk} a_j b_k \left(= \sum _ { j = 1 } ^ { n } \sum _ { k = 1 } ^ { n } \varepsilon_{ijk} a_j b_k\right).
\end{equation}
A useful relation to the Kronecker-Delta is:
\begin{equation} \label{eq:compact_levi}
  \varepsilon _ { i j k } \varepsilon _ { i m n } = \delta _ { j m } \delta _ { k n } - \delta _ { j n } \delta _ { k m }.
\end{equation}
\begin{flushright}\cite{MCGINTY:CONTINUUM}\end{flushright}
\clearpage
\section{Basics}\label{sec:basics}
The first chapters will derive the material point method and necessary extensions for an elastic model. An elastic model is an interesting case displaying much of the needed computations for a typical model. The last chapter \ref{sec:gpgpu} tends to the GPGPU side.
\subsection{Reference and current configuration}\label{sec:ref_corr_config}
Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle. Rather, the continuum assumption holds: Each particle represents a continuous piece of material s.t. a microscopic view does not need to be adopted. A particular body is composed of a set of particles and can adapt different configurations due to changes in shape. These changes are caused by external or internal effects (forces etc.) on it and deform the body over time. \cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}

When modelling solids, changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured. In the material point method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$). This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.

Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the reference and current configuration respectively.  Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$ which relates the reference configuration to the current configuration. Let for simplicity $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations. Then this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
\begin{equation}
  _t\boldsymbol{x} =  {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } )
\end{equation}
If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , this mapping is defined as:
\begin{equation}\label{eq:rigid}
  _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
\end{equation}
The velocity of a material point in the reference configuration can be defined using this mapping
\begin{equation}\label{eq:velocity}
  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial ({^t_0}\phi)}{\partial t}(_0\boldsymbol{x},t),
\end{equation}
and similarly the acceleration is defined
\begin{equation}
  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 ({^t_0}\phi)}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
\end{equation}
It is helpful to abstract away from the reference configuration and think of it as being defined in a different fixed material space. Physically this has the impact of moving with the particle in world space. Commonly known as the Lagrangian form. It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed. The Eulerian description is static: Variables of particles moving by are measured while staying in a fixed position.

These descriptions are different but they will yield the same measurements when related correctly. I.e. both configurations refer variables defined on them to the deformed state but the position, where the 'lookup' of that value happens, is different. In the reference configuration lookup is done at the initial position of the particle. In the current configuration lookup happens at the particle's world position. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
\begin{equation}
  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
\end{equation}
and the (Eulerian) push forward
\begin{equation}\label{eq:push_forward}
  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
\end{equation}
with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.  To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t. an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.

The difficulty in the eulerian formulation becomes apparent when differentiating (due to the chain rule):
\begin{equation} \label{eq:eulerian_general}
  \frac {\partial}{\partial t} {_0f_i } ( {_0\boldsymbol { x }} ,  t  ) =
  \frac { \partial  {_tf_i} } { \partial t } ( {^t_0}\phi ( {_0\boldsymbol {x }} , t ) , t ) + \frac { \partial  {_tf_i }} { \partial {_tx_j} } ( {^t_0}\phi ( {_0\boldsymbol { x }} , t ) , t ) \frac { \partial {^t_0}\phi_j } { \partial t } ( {_0\boldsymbol { x }} , t ).
\end{equation}
Combining this with equation \ref{eq:velocity} and applying the push forward \ref{eq:push_forward} to cancel out mappings leads to the definition referred to as the material derivative in the current configuration:
\begin{equation} \label{eq:material_derivative}
  \frac {  { D } } {   { D }   { t } }   {  {f} } ( _t\boldsymbol { x } ,   { t } ) = \frac { \partial   {  {f} } } { \partial   { t } } ( _t\boldsymbol {x } ,   { t } ) + \frac { \partial   {  {f} } } { \partial x _ {   { j } } } ( _t{\boldsymbol { x }} ,   { t } )  {v} _ {   { j } } ( _t\boldsymbol { x } ,   { t } ).
\end{equation}
The Jacobian of the deformation map $\phi$ is the deformation gradient $\boldsymbol{F}$ and is one of the key components to measure strain for material models:
\begin{equation}
  {^t_0 F_{ij}} ( _0\boldsymbol {x} ,  t  ) = \frac { \partial {^t_0\phi_i} } { \partial _0 x _j } ( _0\boldsymbol { x } , t ) = \frac { \partial _t x _i } { \partial _0 x_j } ( _0\boldsymbol { x } , t ).
\end{equation}
Topology specifies a neighborhood using the open ball concept $_{0,t}B_\epsilon(\boldsymbol{x}) = \{\boldsymbol{y} \in {\Omega^{0,t} | d(\boldsymbol{x},\boldsymbol{y}) < \epsilon}\}$ given a distance measure $d$. $_0B_\epsilon$ becomes the pre-image of $_tB_\epsilon$ under $_0^t\phi$. Intuitively the deformation gradient measures the local deformation of all particles in a small neighborhood $_0B_\epsilon$ to $_tB_\epsilon$. This allows to describe infinitesimal changes in position from the reference to the current configuration
\begin{equation}\label{eq:def_grad_pos}
  d {_tx_i} = {^t_0}F{_{ij}} d{_0x_j}.
\end{equation}
With this quantity in place volume and area changes are calculable. In a typical analytical fashion a coordinate system change $_0x \rightarrow {_tx}$ for a quantity $_0g$ is done using the determinant of the Jacobian matrix. The determinant is given a separate name $^t_0J =\text{det}(^t_0\boldsymbol{F})$. The push forward of $_0g:\Omega^0\rightarrow \mathbb{R}^d$ thus becomes $_tg:\Omega^t\rightarrow \mathbb{R}^d$:
\begin{equation} \label{eq:volume_integral}
  \int _ { {_t\boldsymbol { B } }}  {_tg} ( {_t\boldsymbol { x }} ,t)  d {_t\boldsymbol { x }} = \int _ { { _0\boldsymbol { B }}} {_0 g } ( {_0\boldsymbol{x}},t) {^t_0 J}d {_0\boldsymbol{x}}.
\end{equation}
This can also be achieved by the cross product. A cube spanned by vectors $_0\boldsymbol{x}_i$ ($i=1,2,3$) becomes a parallelepiped in the deformed configuration $d_tV =|{^t_0\boldsymbol{F}}d_0\boldsymbol{x}_0 \cdot({^t_0\boldsymbol{F}}d_0\boldsymbol{x}_1 \times {^t_0\boldsymbol{F}}d_0\boldsymbol{x}_2)| {d_0V}$:
\begin{equation}\label{eq:j}
  d_tV = {^t_0J} d_0V.
\end{equation}
The area change is given by Nanson's Formula. $d\boldsymbol{A}$ is a vector pointing in the direction of the normal of the area and $d_t\boldsymbol{l} ={}^t_0\boldsymbol{F}d_0\boldsymbol{l}$ an arbitrary line element:
$$ d{_0V} = d{_0\boldsymbol{A}}\cdot d{_0\boldsymbol{l}},\quad d{_tV} = d{_t\boldsymbol{A}} \cdot d{_t\boldsymbol{l}}$$
$$ \xRightarrow{\ref{eq:def_grad_pos},\ref{eq:j}} {^t_0J} d_0\boldsymbol{A} \cdot d_0\boldsymbol{l} = d{_t\boldsymbol{A}} \cdot ({^t_0\boldsymbol{F}}d{_0\boldsymbol{l}})$$
\begin{equation}
  \Rightarrow d_t\boldsymbol{A} = {^t_0\boldsymbol{F}}^{-T} {^t_0}J d_0\boldsymbol{A} = {^0_t\boldsymbol{F}}^{T} {^t_0}J d_0\boldsymbol{A}.
\end{equation}
A surface integral may then be transformed to reference configuration by
\begin{equation}\label{eq:surface_integral}
  \int _ { \partial {_0B}  } \boldsymbol{h} ( {_t\boldsymbol{x}} , t ) \cdot d \boldsymbol{A} ( {_t\boldsymbol{x}} ) =  \int _ { \partial {_tB} } \boldsymbol{h} ( {_0\boldsymbol{x}} , t ) \cdot \boldsymbol{F}^{-T}J d \boldsymbol{A} ( {_0\boldsymbol{x}} )
\end{equation}
where $_0\boldsymbol{h} = \boldsymbol{h}(_0\boldsymbol{x},\cdot)$ is the pull back of $_t\boldsymbol{h} = \boldsymbol{h}(_t\boldsymbol{x},\cdot)$. $d_0\boldsymbol{A}$, $d_t\boldsymbol{A}$ point in the direction of the surface normal of $\partial {_0B}(_0\boldsymbol{x})$, $\partial {_tB}(_t\boldsymbol{x})$, respectively.
\cite{MIT:CONTINUUM_MECHANICS}
\cite{MPM:COURSE}
\subsection{Polar and singular value decomposition}\label{sec:svd}
The target is to define strain measures in terms of the deformation gradient: $\boldsymbol{\epsilon}(\boldsymbol{F})$. In equation \ref{eq:rigid} a rigid body movement was introduced. Let $b_i(t) = v_it$ be more generally some translation. A problem arises when calculating the deformation gradient of this equation.
\begin{equation}
  ^t_0F_{ij} = \frac{\partial{_tx_i}}{\partial_0x_j} = \frac{\partial (R_{ik}(t)_0x_k+b_i(t))}{\partial _0x_j} = R_{ik}(t) \delta_{kj} = R_{ij}(t).
\end{equation}
As can be seen, the deformation gradient contains a rigid rotation. For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation (more in chapter \ref{sec:linear_elasticty}). I.e. the deformation gradient has two components a constant rotation and the actual distortion or strain. There are two ways to deal with this:
\begin{enumerate}
  \item  Use a strain measure that cancels out the rotation. An example for this would be the Green-Lagrangian strain tensor with quadratic components:
    \begin{equation}\label{eq:green}

>>>>

\documentclass{beamer}
\usepackage{listings}
\usepackage{xurl}
\urlstyle{sf}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{color}
\usepackage{subfigure}
\usepackage{wrapfig}
\usepackage[english,ngerman]{babel}
\usepackage[backend=biber,style=alphabetic]{biblatex}
\usepackage{mathtools}
\usepackage{subdepth}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{svg}
\usepackage{xcolor,colortbl}

\addbibresource{../mscthesis.bib}

@@

\begin{document}
\begin{frame}
  \begin{enumerate}
    \item \cite{MPM:COURSE}
  \end{enumerate}
\end{frame}
\selectlanguage{english}
\printbibliography
\end{document}
